RBP-J regulates homeostasis and function of circulating Ly6Clo monocytes

Version 1

Abstract

Summary

Notch-RBP-J signaling plays an essential role in maintenance of myeloid homeostasis. However, its role in monocyte cell fate decisions is not fully understood. Here we showed that conditional deletion of transcription factor RBP-J in myeloid cells resulted in marked accumulation of blood Ly6Clo monocytes that highly expressed chemokine receptor CCR2. Bone marrow transplantation and parabiosis experiments revealed a cell intrinsic requirement of RBP-J for controlling blood Ly6CloCCR2hi monocytes. RBP-J-deficient Ly6Clo monocytes exhibited enhanced capacity competing with wildtype counterparts in blood circulation. In accordance with alterations of circulating monocytes, RBP-J deficiency led to markedly increased population of lung tissues with Ly6Clo monocytes and CD16.2+ interstitial macrophages. Furthermore, RBP-J deficiency-associated phenotypes could be genetically corrected by further deleting Ccr2 in myeloid cells. These results demonstrate that RBP-J functions as a crucial regulator of blood Ly6Clo monocytes and thus derived lung-resident myeloid populations, at least in part through regulation of CCR2.

Introduction

Monocytes are integral components of the mononuclear phagocyte system (MPS) that develop from monocyte precursors in the bone marrow1. Several functionally and phenotypically distinct subsets of blood monocytes have been defined on the basis of expression of surface markers2-5. In mice, Ly6Chi monocytes (also called inflammatory or classical monocytes) characterized by Ly6ChiCCR2hiCX3CR1lo exhibit a short half-life, and are recruited to tissue and differentiate into macrophages and dendritic cells6. Ly6Chi monocytes are analogous to CD14+ human monocytes based on gene expression profiling7. A second subtype of monocytes is called Ly6Clo monocytes (also termed patrolling monocytes or non-classical monocytes), which express high level of CX3CR1 and low level of CCR2, equivalent to CD14loCD16+ human monocytes. Ly6Clo monocytes are regarded as blood-resident macrophages that patrol blood vessels and scavenge microparticles attached to the endothelium under physiological conditions, and exhibit a long half-life in the steady-state8-10. Ly6Clo monocytes may exert a protective effect by suppressing atherosclerosis in mice, while high levels of nonclassical monocytes have been associated with more advanced vascular dysfunction and oxidative stress in patients with coronary artery disease11-13. In the inflammatory settings, Ly6Clo monocytes often show anti-inflammatory properties, yet also exhibit a proinflammatory role under certain circumstances5,9,14-18. In addition, Ly6Clo monocytes have demonstrated protective functions in tumorigenesis, such as engulfing tumor materials to prevent cancer metastasis, activating and recruiting NK cells to the lungs19,20. While the functions of Ly6Clo monocytes are complex and sometimes contradictory depending on the animal models and experimental approaches, it is clear that they play a crucial role in both health and disease.

The generation and maintenance of Ly6Clo monocytes are regulated by several transcription factors, including Nr4a1 and CEBPβ21,22. The colony-stimulating factor 1 receptor (CSF1R) signaling pathway is important for the generation and survival of Ly6Clo monocytes, and neutralizing antibodies against CSF1R can reduce their numbers23. Interestingly, recent studies have suggested that different subsets of monocytes may be supported by distinct cellular sources of CSF-1 within bone marrow niches, in which targeted deletion of Csf1 from sinusoidal endothelial cells selectively reduces Ly6Clo monocytes but not Ly6Chi monocytes24. LAIR1 has been shown to be activated by stromal protein Colec12, and LAIR1 deficiency leads to aberrant proliferation and apoptosis in bone marrow non-classical monocytes25. These findings further underscore the complexity of the mechanisms that regulate Ly6Clo monocyte, and suggest that multiple factors are involved in this process.

Recombination signal-binding protein for immunoglobulin kappa J region (RBP-J; also named CSL in humans) is commonly known as the master nuclear mediator of canonical Notch signaling26. In the absence of Notch intracellular domain, RBP-J associates with co-repressor proteins to repress transcription of downstream target genes. In the immune system, the best studied functions of Notch signaling are its roles in regulating the development of lymphocytes, such as T cells and marginal zone B cells26. Notch signaling also has been reported to regulate the differentiation and function of myeloid cells including granulocyte/monocyte progenitors, osteoclasts and dendritic cells 27-30. In osteoclasts, RBP-J represses osteoclastogenesis, particularly under inflammatory conditions30. In dendritic cells, Notch-RBP-J signaling controls the maintenance of CD8- DC28. In addition, Notch2-RBP-J pathway is essential for development of CD8-ESAM+ DC in the spleen and CD103+CD11b+ DC in the lamina propria of the intestine29. However, the role of Notch-RBP-J signaling pathway in regulating monocyte subsets remains elusive. Here, we demonstrated that RBP-J controlled homeostasis of blood Ly6Clo monocytes in a cell intrinsic manner. RBP-J deficiency led to a drastic increase in blood Ly6Clo monocytes, lung Ly6Clo monocytes and CD16.2+ IM. Our results suggest that RBP-J plays a critical role in regulating the population and characteristics of Ly6Clo monocytes in the blood and lung.

Results

RBP-J is essential for maintenance of blood Ly6Clo monocytes

To investigate the role of RBP-J in monocyte subsets, we utilized mice with RBP-J specific deletion in the myeloid cells (Rbpjfl/flLyz2cre/cre mice). Efficient deletion of Rbpj in blood monocytes was confirmed by quantitative real-time PCR (qPCR) (Figure 1 - figure supplement 1A). Flow cytometric analysis revealed that RBP-J deficient mice had a significant increase in the proportion of Ly6Clo monocytes, but not Ly6Chi monocytes, in blood, compared to age-matched control mice with the genotype Rbpj+/+Lyz2cre/cre (Figure 1A). In contrast to circulating monocytes, RBP-J deficient mice exhibited minimal alterations in the percentages of monocyte subsets in bone marrow (BM) and spleen (Figure 1B; Figure 1 - figure supplement 1C). Additionally, RBP-J deficiency in myeloid cells did not appear to have an effect on neutrophils (Figure 1 - figure supplement 1B). Therefore, among circulating myeloid populations, RBP-J selectively controlled the subset of Ly6Clo monocytes.

RBP-J deficient mice display more Ly6Clo blood monocytes.

(A) Blood Ly6Chi and Ly6Clo monocytes in Rbpj+/+Lyz2cre/cre control and Rbpjfl/flLyz2cre/cre mice were determined by flow cytometry analyses (FACS). Representative FACS plots (left) and cumulative data of cell ratio (right) are shown.

(B and C) Representative FACS plots and cumulative data quantitating percentages of BM monocyte subsets (B) and myeloid progenitor cells (C) (CD45+CD11b+Ly6G- CD115+ Ly6Chi monocyte; CD45+CD11b+Ly6G-CD115+Ly6Clo monocyte; MDP, CD45+Lin-CD117+CD115+CD135+ Ly6C-; cMoP, CD45+Lin-CD11b-CD117+CD115+ CD135-Ly6C+). Lin: CD3, B220, Ter119, and Gr-1, CD11b.

(D) Experimental outline for panel (E).

(E) Cumulative data quantitating percentages of sinusoidal monocytes (CD45+) within total BM Ly6Clo monocytes.

Data are pooled from at least 2 independent experiments; n ≥ 4 in each group. Data are shown as mean ± SEM; n.s., not significant; **, P < 0.01 (two-tailed Student’s unpaired t test). Each symbol represents an individual mouse.

BM progenitors that give rise to circulating monocytes are monocyte-dendritic cell progenitors (MDPs) and common monocyte progenitors (cMoPs)21,31-33. Next, we analyzed BM progenitors, and found that the percentages of MDPs and cMoPs were equivalent in control and RBP-J deficient mice (Figure 1C). These results in conjunction with the observations of normal BM monocyte populations suggest that alterations of peripheral blood Ly6Clo monocytes may not originate from BM. Next, we examined whether RBP-J may influence the egress of Ly6Clo monocytes from BM, and measured BM exit rate of Ly6Clo monocytes using in vivo labeling of sinusoidal cells as previously described (Figure 1D)34. The results showed that RBP-J deficiency did not affect egress of Ly6Clo monocytes from BM (Figure 1E). Taken together, these data revealed RBP-J as a critical regulator controlling homeostasis of peripheral blood Ly6Clo monocytes.

RBP-J deficiency does not affect Ly6Chi monocyte conversion or Ly6Clo monocyte survival and proliferation

Next, we aimed to determine whether the increase of blood Ly6Clo monocytes in RBP-J deficient mice was due to decreased cell death or enhanced proliferation. We first stained monocytes with Annexin V and 7-AAD to identify apoptotic cells and observed comparable percentages of apoptotic blood Ly6Clo monocytes in control and RBP-J deficient mice (Figure 2A). Given the crucial role of Nr4a1 in the survival of Ly6Clo monocytes21, we detected the expression of Nr4a1, which was similar in Ly6Clo monocytes from control and RBP-J deficient mice (Figure 2 - figure supplement 1A). We then assessed the proliferative capacity by analyzing Ki-67 expression as well as in vivo EdU incorporation. Ki-67 levels in blood monocytes displayed no differences between control and RBP-J deficient mice (Figure 2B). The percentage of EdU+ monocytes did not significantly differ between control and RBP-J deficient mice at the indicated time points, implying that the turnover of Ly6Clo monocytes was normal in RBP-J deficient mice (Figure 2C). Fluorescent latex beads as particulate tracers could be phagocytosed by monocytes after intravenous injection, and stably label Ly6Clo monocytes35. Thus, we intravenously injected fluorescent latex beads into control and RBP-J deficient mice to track circulating monocytes (Figure 2D). By day 7 post injection, only Ly6Clo monocytes were latex+ as previously reported35, whereas control and RBP-J deficient mice groups presented a similar frequency of latex+ monocytes (Figure 2E). Together, these data indicated that RBP-J did not influence monocyte survival and proliferation.

Monocyte subsets in RBP-J deficient mice display normal cell death.

(A) Representative FACS plots of monocyte subsets in blood stained with 7-AAD and Annexin V.

(B) FACS analysis of Ki-67 expression in Rbpj+/+Lyz2cre/cre control and Rbpjfl/flLyz2cre/creblood monocyte subsets. Black lines represent control mice and red lines represent RBP-J deficient mice.

(C) Analysis of time course of EdU incorporation of monocyte subsets in BM and blood after a single 1mg EdU pulsing. The percentages of EdU+ cells among the indicated monocyte subsets are shown.

(D) Experimental outline for panel (E).

(E) Analysis of time course of latex beads incorporation of monocyte subsets in blood after Latex beads injection. The percentages of latex+ cells among the indicated monocyte subsets are shown.

(F) Cartoon depicting the adoptive transfer. BM GFP+Ly6Chi monocytes were sorted from Lyz2cre/creCx3cr1gfp/+or Rbpjfl/flLyz2cre/creCx3cr1gfp/+mice and transferred into Ccr2RFP/RFP recipient mice. Sixty hours after transfer, cell fate was analyzed.

(G) Representative FACS plots are shown in the left panel, and the frequencies of GFP+ Ly6Clo monocytes within total Ly6Clo monocytes are shown in the right panel.

Data are pooled from 2 independent experiments (G); n ≥ 2 in each group (C, E and G). Data are shown as mean ± SEM; n.s., not significant; (two-tailed Student’s unpaired t test). Each symbol represents an individual mouse (E and G).

Previous studies have shown that Ly6Clo monocytes are observed in recipient mice following the adoptive transfer of Ly6Chi monocytes, indicating that Ly6Chi monocytes can convert into Ly6Clo monocytes10,31. We next wished to evaluate whether conversion of Ly6Chi monocytes was regulated by RBP-J by isolating bone marrow GFP+Ly6Chi monocytes from Lyz2cre/creCx3cr1gfp/+control or Rbpjfl/flLyz2cre/creCx3cr1gfp/+mice and adoptively transferring them into Ccr2RFP/RFP recipients (Figure 2F). Sixty hours after transfer, a subset of cells from donors were converted into Ly6Clo monocytes (Figure 2 - figure supplement 1B), and equal percentages of Ly6Clo monocytes were derived from control and Rbpjfl/flLyz2cre/creCx3cr1gfp/+donors (Figure 2G). Thus, the conversion of Ly6Chi monocytes into Ly6Clo monocytes was not affected by RBP-J deficiency.

RBP-J regulates blood Ly6Clo monocytes in a cell intrinsic manner

We next wondered whether the increase of Ly6Clo monocytes in RBP-J deficient mice was bone marrow-derived and cell intrinsic. We performed bone marrow transplantation by engrafting lethally irradiated mice with a 1:4 mixture of Rbpj+/+Lyz2cre/cre control and Rbpjfl/flLyz2cre/cre BM cells (CD45.2) and Cx3cr1gfp/+ BM cells (CD45.1) (Figure 3A). Eight weeks after transplantation, we analyzed the frequencies of CD45.2+ donor cells in the bone marrow and blood of recipient mice and found that the frequencies of CD45.2+ cells within total cells were similar to the mixture ratio of bone marrow cells for Ly6Chi monocytes and neutrophils in both BM and blood (Figure 3B; Figure 3 - figure supplement 1A-B). Specifically, more blood Ly6Clo monocytes were derived from RBP-J deficient donors than control donors (Figure 3B), reflecting that the contribution of RBP-J deficient cells in blood Ly6Clo cells was significantly higher than that of control cells. These results implied that RBP-J deficient BM cells were highly efficient in generating blood Ly6Clo monocytes.

The role of RBP-J in blood Ly6Clo monocytes is cell intrinsic.

(A) Cartoon depicting the bone marrow transplantation. BM recipient 6-week-old male C57BL/6 mice (CD45.2) were lethally irradiated. BM cells from donor mice (Rbpj+/+Lyz2cre/cre or Rbpjfl/flLyz2cre/cre and CD45.1, Cx3cr1gfp/+) were collected and transferred into recipient mice. Mice were used after 8 weeks of bone marrow reconstitution.

(B) Representative FACS plots (left) and cumulative data (right) quantitating the frequency of Rbpj+/+Lyz2cre/cre and Rbpjfl/flLyz2cre/cre donor cells among Ly6Chi and Ly6Clo monocytes in the blood of recipient mice.

(C) Cartoon depicting the generation of Rbpj+/+Lyz2cre/cre control or Rbpjfl/flLyz2cre/cre and Cx3cr1gfp/+ parabiotic pairs.

(D) Representative FACS plots (left) and cumulative data (right) quantitating percentages of monocyte subsets derived from control or RBP-J deficient mice in control or RBP-J deficient mice.

(E) Representative FACS plots (left) and cumulative data (right) quantitating percentages of monocyte subsets derived from control or RBP-J deficient mice in Cx3cr1gfp/+ mice.

Data are pooled from at least 2 independent experiments; n ≥ 6 in each group. Data are shown as mean ± SEM; n.s., not significant; ***, P < 0.001; ****, P < 0.0001 (two-tailed Student’s unpaired t test). Each symbol represents an individual mouse. SSC-A, side scatter area.

Given that RBP-J deficient mice had more Ly6Clo monocytes in blood than did control mice, we performed parabiosis experiments, a surgical union of two organisms, which allowed parabiotic mice to share their blood circulation. The contribution of circulating cells from one animal in another can be estimated by measuring the percentage of blood cells that originated from each animal36. We joined a CD45.1, Cx3cr1gfp/+ mouse with an age and sex-matched Rbpj+/+Lyz2cre/cre control or Rbpjfl/flLyz2cre/cre mouse (CD45.2) (Figure 3C). Four weeks after the procedure, about 50% of B cells and T cells in parabiotic mice displayed efficient exchange of their circulation (Figure 3 - figure supplement 1C). As expected, RBP-J deficient mice exhibited higher percentages of Ly6Clo monocytes than control animals (Figure 3D). Intriguingly, in the parabiotic Cx3cr1gfp/+mice, RBP-J deficient cells still constituted significantly higher proportion of circulating Ly6Clo, but not Ly6Chi monocytes, than RBP-J sufficient counterparts as indicated by the percentages of the GFP-negative population (Figure 3E). These results implicated that the enhanced ability of Ly6Clo monocytes to circulate in the peripheral blood as a result of RBP-J deficiency was cell intrinsic.

RBP-J regulates phenotypical marker genes in blood Ly6Clo monocytes

To further study the consequences of RBP-J loss of function in blood Ly6Clo monocytes, we performed gene expression profiling by RNA-seq, which revealed that Ly6Clo monocytes in RBP-J deficient mice exhibited low expression of Itgax but high expression of Ccr2, in comparison to those in control mice (Figure 4A). These differential expression patterns of phenotypic marker genes were also confirmed by qPCR and flow cytometry analysis (Figure 4B). However, Ly6Chi blood monocytes and BM monocytes displayed normal levels of CD11c and CCR2 (Figure 4B; Figure 4 - figure supplement 1A). Moreover, principal component analysis showed distinct expression pattern of monocytes between control and RBP-J deficient mice (Figure 4C). To determine whether the regulation of phenotypic markers by RBP-J in blood Ly6Clo monocytes was cell-intrinsic, we performed BM transplantation as described above (Figure 3A). The findings indicated that Ly6Clo monocytes derived from RBP-J deficient BM cells expressed high levels of CCR2 but low levels of CD11c, in comparison to those derived from control BM cells, whereas no significant changes were observed in blood Ly6Chi and BM monocytes, as expected (Figure 4D; Figure 4 - figure supplement 1B). These results collectively suggested that RBP-J regulated the expression of CCR2/CD11c in a cell-intrinsic manner.

Phenotypic markers in Ly6Clo monocytes are changed in RBP-J deficient mice.

(A) Heatmap of RNA-seq dataset showing the top 20 downregulated and upregulated genes in blood Ly6Clo monocytes from Rbpjfl/flLyz2cre/cre versus Rbpj+/+Lyz2cre/cre control mice. Blue and red font indicates downregulated and upregulated genes in RBP-J deficient Ly6Clo monocytes respectively.

(B) Representative FACS plots, cumulative MFI, and qPCR analysis of CCR2/CD11c expression in control and RBP-J deficient blood monocyte subsets. Shaded curves represent isotype control, black lines represent control mice and red lines represent RBP-J deficient mice.

(C) PCA of indicated cell types.

(D) Representative FACS plots and cumulative MFI of CCR2/CD11c expression in blood monocyte subsets derived from control or RBP-J deficient mice. Black lines represent control mice, and red lines represent RBP-J deficient mice.

Data are pooled from 2 independent experiments (B and D); n ≥ 6 in each group. Data are shown as mean ± SEM; n.s., not significant; **, P < 0.01; ****, P < 0.0001 (two-tailed Student’s unpaired t test). Each symbol represents an individual mouse.

RBP-J-mediated control of blood Ly6Clo monocytes is CCR2 dependent

Given that RBP-J deficiency led to enhanced CCR2 expression in Ly6Clo monocytes and that CCR2 is essential for monocyte functionality under various inflammatory and non-inflammatory conditions37, we wished to genetically investigate the role of CCR2 in the RBP-J deficient background. We generated RBP-J/CCR2 double deficient (DKO) mice with the genotype Rbpjfl/flLyz2cre/creCcr2RFP/RFPwith Lyz2cre/creCcr2RFP/+ and Rbpjfl/flLyz2cre/creCcr2RFP/+mice serving as controls. As expected, the expression of CCR2 in Ly6Clo monocytes was reduced in DKO mice compared to RBP-J deficient mice (Figure 5A), confirming the successful deletion of the Ccr2 gene. DKO mice showed a lower percentage of both Ly6Chi and Ly6Clo monocytes than RBP-J deficient mice (Figure 5B-C). Notably, the percentage of Ly6Clo monocytes in DKO mice was comparable to that observed in the control mice (Figure 5B-C), implicating that deletion of CCR2 corrected RBP-J deficiency-associated phenotype of increased Ly6Clo monocytes. Whereas, both RBP-J deficient and DKO mice exhibited lower expression levels of CD11c in their Ly6Clo monocyte than control mice, suggesting that the regulation of CD11c expression was independent of CCR2 (Figure 5D). These results suggested that RBP-J regulated Ly6Clo monocytes, at least in part, through CCR2.

Blood Ly6Clo monocytes are decreased in DKO mice.

(A) Representative FACS plots and cumulative MFI of CCR2 expression in Lyz2cre/creCcr2RFP/+ control, Rbpjfl/flLyz2cre/creCcr2RFP/+ and Rbpjfl/flLyz2cre/creCcr2RFP/RFP (DKO) blood Ly6Chi and Ly6Clo monocytes are shown.

(B and C) Blood monocyte subsets in control, RBP-J deficient and DKO mice were determined by FACS. Representative FACS plots (B) and cumulative data of cell ratio

(C) are shown.

(D) Representative FACS plots and cumulative MFI of CD11c expression in control, RBP-J deficient and DKO blood Ly6Chi and Ly6Clo monocytes are shown.

Data are pooled from at least 2 independent experiments; n ≥ 5 in each group. Data are shown as mean ± SEM; n.s., not significant; **, P < 0.01; ***, P < 0.001; ****, P < 0.0001 (two-tailed Student’s unpaired t test). Each symbol represents an individual mouse.

Lung Ly6Clo monocytes are accumulated in RBP-J deficient mice

In mice, alveolar macrophages (AM) are maintained by local self-renewal, and arise from fetal liver derived precursors, whereas interstitial macrophages (IM) are probably originated from monocytes 38-40. Joey et al identified 3 subpopulations of IM according to expression of CD206 and CD16.2 and Ly6Clo monocytes are proposed to give rise to CD64+CD16.2+ IM41. We next compared the populations of lung monocytes and IM between control and RBP-J deficient mice. The conditional deletion of RBP-J resulted in a significant increase in the absolute and relative numbers of Ly6Clo monocytes and CD16.2+ IM, while the numbers of Ly6Chi monocytes, CD206- IM, and CD206+ IM remained unchanged (Figure 6A-B). To confirm these findings, lung sections from Lyz2cre/creCx3cr1gfp/+ control and Rbpjfl/flLyz2cre/creCx3cr1gfp/+ mice were stained with an anti-GFP antibody, which revealed a significant increase in the number of GFP+ cells in RBP-J deficient mice (Figure 6C). In addition, we evaluated the proliferative capacity of Ly6Clo monocytes and CD16.2+ IM, but did not observe enhanced EdU incorporation in Ly6Clo monocytes and CD16.2+ IM from RBP-J deficient mice (Figure 6 - figure supplement 1A-B). These findings suggested that the augmented population of these cells might not be resulted from increased in situ proliferation. Previous reports have indicated that LPS can increase the population of IM in a CCR2-dependent manner40. We thus challenged mice with LPS, and analyzed monocytes and IM at day 0 and day 4. LPS exposure induced an increase in numbers of Ly6Clo monocytes, CD16.2+ IM and CD206- IM compared to baseline both in control and RBP-J deficient mice (Figure 6D-E). There was a trend toward higher Ly6Chi monocyte after LPS treatment, although this observation was not statistically significant. Of note, RBP-J deficient mice exhibited robustly elevated numbers of Ly6Clo monocytes and CD16.2+ IM compared with control mice after LPS treatment (Figure 6D-E), suggesting that RBP-J deficient Ly6Clo monocytes were recruited to inflamed tissue in large numbers and differentiated into CD16.2+ IM.

RBP-J deficient mice exhibit more lung Ly6Clo monocytes and CD16.2+ IM.

(A and B) Indicated populations in the lungs of Rbpj+/+Lyz2cre/cre and Rbpjfl/flLyz2cre/cre mice were determined by FACS. Representative FACS plots (A) and cumulative data of cell ratio and absolute numbers (B) are shown.

(C) Immunofluorescence staining for GFP+ cells in the lung from Lyz2cre/creCx3cr1gfp/+and Rbpjfl/flLyz2cre/creCx3cr1gfp/+mice (CX3CR1 [green]; DAPI [blue]). Scale bars represent 50 µm.

(D and E) Rbpj+/+Lyz2cre/cre and Rbpjfl/flLyz2cre/cre mice were instilled intranasally with PBS or PBS containing LPS, and lungs were harvested at the indicated time points. Representative FACS plots (D) and cumulative data of cell ratio and absolute numbers

(E) are shown.

Data are pooled from at least 2 independent experiments; n ≥ 4 in each group. Data are shown as mean ± SEM; n.s., not significant; *, P < 0.05; **, P < 0.01; ****, P < 0.0001 (two-tailed Student’s unpaired t test). Each symbol represents an individual mouse.

To investigate whether the elevated levels of lung monocytes and CD16.2+ IM in RBP-J deficient mice were derived from blood Ly6Clo monocytes, we examined monocytes and IM in Lyz2cre/creCcr2RFP/+control, Lyz2cre/creCcr2RFP/RFP, Rbpjfl/flLyz2cre/creCcr2RFP/+, and Rbpjfl/flLyz2cre/creCcr2RFP/RFP(DKO) mice. The deletion of CCR2 led to a decrease in Ly6Chi monocytes, and the Ly6Clo monocytes in DKO mice were reduced to a level similar to that of control mice (Figure 7A, B). While the DKO mice had more CD16.2+ IM compared to control mice, these cells were severely reduced compared to RBP-J deficient mice (Figure 7A, C). The above data supported the notion that increased lung Ly6Clo monocytes and CD16.2+ IM in RBP-J deficient mice were derived from Ly6Clo blood monocytes. In summary, these results suggested that in RBP-J deficient mice, recruitment of blood Ly6Clo monocytes to the lung was markedly facilitated by increased cell number as well as heightened expression of CCR2, leading to the increase of lung Ly6Clo monocytes and CD16.2+ IM.

DKO mice lack lung Ly6Clo monocytes and CD16.2+ IM.

(A) Representative FACS plots of lung monocyte and IM subsets in Lyz2cre/creCcr2RFP/+, Rbpjfl/flLyz2cre/creCcr2RFP/+, Lyz2cre/creCcr2RFP/RFP and Rbpjfl/flLyz2cre/creCcr2RFP/RFP mice.

(B and C) Cumulative data of cell ratio and absolute numbers of monocyte (B) and IM (C) subsets.

Data are pooled from 2 independent experiments; n ≥ 4 in each group. Data are shown as mean ± SEM; n.s., not significant; *, P < 0.05; **, P < 0.01; ***, P < 0.001; ****, P < 0.0001 (two-tailed Student’s unpaired t test). Each symbol represents an individual mouse.

Discussion

In this study, we identified RBP-J as a pivotal factor in regulating blood Ly6Clo monocyte cell fate. Our results showed that mice with conditional deletion of RBP-J in myeloid cells exhibited a robust increase in blood Ly6Clo monocytes, and subsequently accumulated lung Ly6Clo monocytes and CD16.2+ IM under steady-state conditions. Bone marrow transplantation experiments in which RBP-J deficient cells were transplanted into recipient mice as well as the parabiosis experiment showed similar increase in blood Ly6Clo monocytes, demonstrating that RBP-J was an intrinsic factor required for the regulation of Ly6Clo monocytes. Further analysis revealed that RBP-J regulated the expression of CCR2 and CD11c on the surface of Ly6Clo monocytes. Moreover, the phenotype of elevated blood Ly6Clo monocytes and their progeny was further ameliorated by deleting Ccr2 in an RBP-J deficient background.

Notch-RBP-J signaling have been shown to play a role in regulating functional polarization and activation of macrophages, and regulated formation of Kupffer cells and macrophage differentiation from Ly6Chi monocytes in ischemia42-48. At steady-state, interaction of DLL1 with Notch 2 regulates conversion of Ly6Chi monocytes into Ly6Clo monocytes in special niches of the BM and spleen49. Under inflammatory conditions, Notch2 and TLR7 pathways independently and synergistically promote conversion of Ly6Chi monocytes into Ly6Clo monocytes50. However, in our study, the percentage of blood Ly6Clo monocytes increased in RBP-J deficient mice. These results somewhat differ from what was observed in mice with conditional deletion of Notch249. Actually, 4 members of the Notch family have been identified in mammals26. Thus, Notch receptors seem to be non-redundant in regulating monocyte cell fate, and distinct Notch receptors are likely to act in a concerted manner to coordinate the monocyte differentiation program.

Colony-stimulating factor 1 receptor (CSF1R) signal, CX3CR1, C/EBPβ and Nr4a1 have been suggested to be involved in the generation and survival of Ly6Clo monocytes. The lifespan of Ly6Clo monocytes is acutely shortened after blockade of CSF1R, and the numbers of these cells are reduced, and the percentage of dead cells increased in mice with endothelial cell-specific depletion of Csf123,24. CX3CR1 and C/EBPβ-knockout mice show accelerated death of Ly6Clo monocyte, display the decreased level of Ly6Clo monocytes22,51. Nr4a1 knockout mice had significantly fewer monocytes in blood, bone marrow, and spleen, due to differentiation deficiency in MDP and accelerated apoptosis of Ly6Clo monocytes in bone marrow21. RBP-J deficient mice had normal MDP and Ly6Clo monocytes, expressed normal level of Nr4a1 and Ki-67, exhibited normal half-life and turnover rate, which suggested that RBP-J may not regulate Ly6Clo monocytes through these factors. Further studies are required to elucidate the precise molecular mechanisms of RBP-J in Ly6Clo monocytes under steady state.

At the steady state, Ly6Clo monocytes patrol endothelium of blood vessel. After R848 induced endothelial injury, Ly6Clo monocytes recruit neutrophils to mediate focal endothelial necrosis and fuel vascular inflammation, and subsequently, the Ly6Clo monocytes remove cellular debris9,52. In response to Listeria monocytogenes infection, Ly6Clo monocytes can extravasate but do not give rise to macrophages or DCs8. Additional evidence shows that Ly6Clo monocytes do not merely act as luminal blood macrophages. In lung, exposure to unmethylated CpG DNA expands CD16.2+ IM, which originate from Ly6Clo monocytes and spontaneously produce IL-10, thereby preventing allergic inflammation41. A recent study shows that Ly6Clo monocytes give rise to CD9+ macrophages, which provide an intracellular replication niche for Salmonella Typhimurium in spleen, and Ly6Clo-depleted mice are more resistant to Salmonella Typhimurium infection, suggesting that Ly6Clo monocytes exert certain functions in systemic infection53. Our data provide evidence for RBP-J in the function of blood Ly6Clo monocytes, as RBP-J deficient Ly6Clo monocytes exhibited enhanced competition in blood circulation, as well as gave rise to increased numbers of lung CD16.2+ IM. In summary, we showed that RBP-J acted as a fundamental regulator of the maintenance of blood Ly6Clo monocytes and their descendent, at least in part through regulation of CCR2. These results provide insights into understanding the mechanisms that regulate monocyte homeostasis and function.

eLife Assessment

This study presents a valuable examination into the role Notch-RBP-J signaling in regulating monocyte subset homeostasis. The data were collected and analyzed using solid and validated methodology and can be used as a starting point for exploring the mechanisms involved in RBP-J signaling in non-classical monocytes.

Reviewer #1 (Public Review):

Kou and Kang et al. investigated the role of Notch-RBP-J signaling in regulating monocyte homeostasis. Specifically, they examined how a conditional knockout of Rbpj expression in monocytes through a Rbpjfl/fl Lyz2cre/cre mouse affects the homeostasis of Ly6Chi versus Ly6Clo monocytes. They discovered that Rbpj deficiency did not affect the percentage of Ly6Chi monocytes but instead, led to an accumulation of Ly6Clo monocytes in the peripheral blood. Using a comprehensive number of in vivo techniques to investigate the origin of this increase, the authors revealed that the accumulation of Rbpj deficient Ly6Clo monocytes was not due to an increase in bone marrow egress and that this defect was cell intrinsic. However, EdU-labelling and apoptosis assays revealed that this defect was not due to an increase in proliferation nor conversion of Ly6Chi to Ly6Clo monocytes. Interestingly, it was revealed that Rbpj deficient Ly6Clo monocytes had increased expression of CCR2 and ablation of CCR2 expression reversed the accumulation of these cells in the periphery. Lastly, they discovered that Rbpj deficiency also led to downstream effects such as an accumulation of Ly6Clo monocytes in the lung tissue and increased CD16.2+ interstitial macrophages, presumably due to dysregulated monocyte differentiation and function.

Their findings are interesting and highlight a previously unexplored mechanistic link between Notch-RBP-J signaling and CCR2 expression in monocyte homeostasis, providing further insight into the distinct pathways that regulate Ly6Chi vs Ly6Clo monocyte subsets individually.

The conclusions of this paper are mostly well substantiated from the experimental data. The strengths of this paper include the use of multiple conditional genetic knock out mouse models to explore their hypothesis and the use of sophisticated in vivo techniques to study the major mechanisms involved in monocyte homeostasis.

Reviewer #2 (Public Review):

The authors provide compelling data to demonstrate that the Notch-related transcription factor RBP-J can influence the number of circulating and recruited monocytes. The authors first delete the Rbpj gene in the myeloid lineage (Lyz2) and show that, as a proportion, only Ly6Clo monocytes are increased in the blood. The authors then attempted to identify why these cells were increased but ruled out proliferation or reduced apoptosis. Next, they investigated the gene signature of Rbpj null monocytes using RNA-sequencing and identified elevated Ccr2 as a defining feature. Crossing the Rbpj null mice to Ccr2 null mice showed reduced numbers of Ly6Clo monocytes compared with Rbpj null alone. Finally, the authors identify that an increased burden of blood Ly6Clo monocytes is correlated with increased lung recruitment and expansion of lung interstitial macrophages.

The main conclusion of the authors, that there is a 'cell intrinsic requirement of RBP-J for controlling blood Ly6CloCCR2hi monocytes' is strongly supported by the data. However, other claims and aspects of the study require clarification and further analysis of the data generated.

Strengths

The paper is well written and structured logically. The major strength of this study is the multiple technically challenging methods used to reinforce the main finding (e.g. parabiosis, adoptive transfer). The finding reinforces the fact that we still know little about how immune cell subsets are maintained in situ, and this study opens the way for novel future work. Importantly, the authors have generated an RNA-sequencing dataset that will prove invaluable for identifying the mechanism - they have promised public access to this data via GEO.

Weaknesses - The main weakness of the study, is that although the main result is solidly supported, as written it is mostly descriptive in nature. For instance, there is no given mechanism by which RBP-J increases Ly6Clo monocytes. The authors conclude this is dependent on CCR2, however CCR2 deletion has a global effect on monocyte numbers and importantly in this study, it does not remove the Ly6Clo bias of cell proportions, if anything it seems to enhance the difference between the ly6C low and high populations in Rbpj null mice (figure 5C). This oversight in data interpretation likely occurred because this experiment is missing a potentially important control (Lyz2cre/cre Ccr2RFP/RFP or RBP-J variations). In general, there seemed to be a focus on the Ly6C low cells, where the mechanism may be more identifiable in their precursors - likely the Ly6C high monocytes.

Other specific weaknesses were identified:

1. The confirmation of knockout in supplemental figure 1A shows only a two third knockdown when this should be almost totally gone. Perhaps poor primer design, cell sorting error or low Cre penetrance is to blame, but this is below the standard one would expect from a knockout.

2. Many figures (e.g. 1A) only show proportional data (%) when the addition of cell numbers would also be informative

3. Many figures only have an n of 1 or 2 (e.g. 2B, 2C)

4. Sometimes strong statements were based on the lack of statistical significance, when more n number could have changed the interpretation (e.g. 2G, 3E)

5. There is incomplete analysis (e.g. Network analysis) and interpretation of RNA-sequencing results (figure 4), the difference between the genotypes in both monocyte subsets would provide a more complete picture and potentially reveal mechanisms

6. The experiments in Figures 5 and 7 are missing a control (Lyz2cre/cre Ccr2RFP/RFP or the Rbpj+/+ versions) and may have been misinterpreted. For example if the control (RBP-J WT, CCR2 KO) was used then it would almost certainly show falling Ly6C low numbers compared to RBP-J WT CCR2 WT, but RBP-J KO CCR2 KO would still have more Ly6c low monocytes than RBP-J WT, CCR2 KO - meaning that the RBP-J function is independent of CCR2. I.e. Ly6c low numbers are mostly dependent on CCR2 but this is irrespective of RBP-J.

7. Figure 6 was difficult to interpret because of the lack of shown gating strategy. This reviewer assumes that alveolar macrophages were gated out of analysis

8. The statements around Figure 7 are not completely supported by the evidence, i) a significant proportion of CD16.2+ cells were CCR2 independent and therefore potentially not all recently derived from monocytes, and ii) there is nothing to suggest that the source was not Ly6C high monocytes that differentiated - the manuscript in general seems to miss the point that the source of the Ly6C low cells is almost certainly the Ly6C high monocytes - which further emphasises the importance of both cells in the sequencing analysis

9. The authors did not refer to or cite a similar 2020 study that also investigated myeloid deletion of Rbpj (Qin et al. 2020 - https://doi.org/10.1096/fj.201903086RR). Qin et al identified that Ly6Clo alveolar macrophages were decreased in this model - it is intriguing to synthesise these two studies and hypothesise that the ly6c low monocytes steal the lung niche, but this was not discussed

Reviewer #3 (Public Review):

In this study, the authors investigate the role of the Notch signalling regulator RBP-J on Ly6Clow monocyte biology starting with the observation that RBP-J-deficient mice have increased circulating Ly6low monocytes. Using myeloid specific conditional mouse models, the authors investigate how RBP-J deficiency effects circulating monocytes and lung interstitial macrophages.

A major strength of this study is that it describes RBP-J as a novel, critical factor regulating Ly6Clow monocyte cell frequency in the blood. The authors demonstrate that RBP-J deficiency leads to increased Ly6Clow monocytes in the blood and lung and CD16.2+ interstitial macrophages in steady state. The authors use a number of different techniques to confirm this finding including bone marrow transplantation experiments and parabiosis.

There are several critical weaknesses that need to be assessed to improve the manuscript, in summary the data presented in the current manuscript are highly descriptive and without mechanistic insight. The inclusion of more mechanistic insight would greatly improve the manuscript.

The authors begin to explore the potential mechanism underlying why Ly6Clow monocytes are increased in the absence of RBP-J - is it through increased survival, increased conversion from Ly6C+ monocytes, increased proliferation or increased egress from the bone marrow. The majority of the data they present here is negative. Whilst I applaud the authors for including negative data, I think that their exploration into how RBP-J leads to increased monocytes does not go far enough and it is critical to understand the mechanism by which RBP-J increases circulating monocytes. Low n-numbers in multiple figures mean that the claims made are not fully supported.

The current title of the paper "RBP-J regulates homeostasis and function of circulating Ly6Clo monocytes" does not fully reflect the manuscript in its current form - there is no exploration of Ly6Clow monocyte functionality in the paper as it stands.

Given that targeting monocytes and macrophages in a range of inflammatory diseases is an attractive yet elusive therapeutic option, understanding the underlying biology that regulates monocyte biology are critically important. This manuscript has the potential to add to our current knowledge of how Ly6Clow monocyte biology is regulated and potentially opens novel avenues for preferentially enhancing Ly6Clow monocytes without influencing Ly6C+ monocytes. This is an attractive proposition for many inflammatory conditions however, considerably more in-depth analysis is required to understand the role of RBP-J in monocyte biology.

Author Response

The following is the authors’ response to the original reviews.

We thank the reviewers for their time and insightful and constructive comments. We are pleased that reviewers found this study “opens the way for novel future work” and the findings “interesting”. We have experimentally addressed the points raised by the reviewers and have substantially revised the manuscript by modifying 30 figures panels. The reviewers’ points are specifically addressed below.

1. The authors concluded that an accumulation of Ly6Clo monocytes occurred in the Rbpjfl/fl Lyz2cre/cre mouse by examining the percentage of cells among CD45+ cells in Figure 1. It would be helpful if the authors could give an account of the total cell count numbers of monocyte subsets per ml of blood and in the bone marrow to give the readers a better idea of the extent of increase as cell percentages among CD45+ cells may be influenced by the number of other immune subsets.

We thank the reviewer for raising these points. In this research, we crossed Rbpjfl/fl mice with Lyz2-Cre mice carrying the Cre recombinase inserted in the Lysozyme-M (Lyz2) gene locus results in the selective deletion of RBP-J in myeloid cells, such as monocytes, macrophages and granulocytes. We then proceeded to examine the neutrophil levels in the bone marrow and blood. The percentage of neutrophils observed was found to be similar to that of control mice, which was in line with the findings reported in the literature (Metzemaekers et al. 2020). Furthermore, the proportion of Ly6Chi monocytes in RBP-J deficient mice was found to be similar to that of control mice, which is consistent with the literature (Ginhoux et al. 2014). Based on these results, we thought that the changes observed in the proportion of Ly6Clo monocytes could reliably indicate the alterations occurring in Ly6Clo monocytes within the Rbpjfl/flLyz2cre/cre mice.

1. The authors demonstrated no significant differences in bone marrow progenitor and monocyte numbers, therefore concluding that monocyte egress from the bone marrow did not contribute to the increase in Ly6Clo monocyte numbers in the blood (Figure 1B-D). As it is unclear what is the exact cell number increase in the blood, the changes in bone marrow monocyte numbers might be too small to be reflected in their percentage calculations. In light that CCR2 was also found to play a role in Ly6Clo monocyte homeostasis in Rbpjfl/fl Lyz2cre/cre mice, could the authors demonstrate if Rbpj-deficient Ly6Clo monocytes might be more responsive to CCL2 through transwell experiments? This would also provide readers a more in-depth mechanism of how an increase in CCR2 on Rbpj-deficient Ly6Clo monocytes leads to their accumulation in the periphery.

The experimental results regarding the proportion of monocytes and precursor cells in the bone marrow were derived from multiple experiments. The data obtained from individual experiments as well as the final integrated data did not reveal significant differences between the control mice and Rbpjfl/flLyz2cre/cre mice. Therefore, we believed that even if there were small changes in cell numbers, these differences could still be reflected through alterations in their proportions. We attempted transwell experiments, but unfortunately, they were not technically successful. Nearly all sorted Ly6Clo monocytes attached to the transwell membrane, making it challenging to draw a conclusion regarding the responsiveness of RBP-J deficient Ly6Clo monocytes to CCL2.

1. In the parabiosis experiment conducted in Figure 3C-E, the authors provide conclusive evidence that the accumulation of Rbpj-deficient Ly6Clo monocytes was cell intrinsic as Rbpj-deficient Ly6Clo monocytes continued to accumulate in the blood of control counterparts. Monocytes have also been shown to accumulate in the spleen and re-enter or home back to the bone marrow. Assessing if there is a change in monocyte homing abilities in Rbpj-deficient Ly6Clo monocytes by examining their numbers in the spleen and bone marrow of control parabiotic mice would substantiate their claims that the defect was cell intrinsic and provide further understanding for the readers of why Rbpj-deficient Ly6Clo monocytes accumulate in the blood.

We thank the reviewer for bringing out this interesting point. We also analyzed the proportions of GFP- Ly6Chi monocytes and Ly6Clo monocytes in the bone marrow of parabiotic mice. The experimental results revealed that there were no significant differences in the proportion of GFP- monocytes between the control mice and the KO animals (see the figure A below). We also detected the expression of CXCR4 in bone marrow Ly6Clo monocytes. Rbpjfl/flLyz2cre/cre mice exhibited normal expression of CXCR4 (see Author response image 1 below), which participates in the homing of classical and nonclassical monocytes to bone marrow and spleen monocyte reservoirs (Chong et al. 2016). The homing abilities of RBP-J deficient Ly6Clo monocytes may not have changed.

1. Authors should provide cell counts for Figure 5B to demonstrate the extent CCR2 depletion affects the number of Ly6Clo monocytes in Rbpjfl/fl Lyz2cre/cre mice as explained in point 1.

As mentioned before, we believed that the proportion of circulating monocytes could, to some extent, provide evidence of the impact of CCR2 deficiency on Ly6Clo monocytes.

Reviewer #2

1. The confirmation of knockout in supplemental figure 1A shows only a two third knockdown when this should be almost totally gone. Perhaps poor primer design, cell sorting error or low Cre penetrance is to blame, but this is below the standard one would expect from a knockout.

Kang et al (PMID: 31944217) evaluated the knockout efficiency of Rbpj in sorted colonic macrophages of Rbp-jfl/flLyz2cre/cre mice using qPCR and immunoblotting. The qPCR result indicated a two-third knockdown, while the immunoblotting results demonstrated efficient deletion of RBP-J protein in Rbp-jfl/flLyz2cre/cre mice. As pointed out by the reviewer, the observed two-third knockdown, which is lower than the expected complete knockout, may be attributed to primer design.

1. Many figures (e.g. 1A) only show proportional data (%) when the addition of cell numbers would also be informative

We appreciate the reviewer for bringing up these points. Indeed, multiple articles studying monocytes only show changes in cell proportions. As mentioned above, we believed that analyzing the proportion of circulating monocytes could offer valuable evidence of the influence of RBP-J deficiency on Ly6Clo monocytes.

1. Many figures only have an n of 1 or 2 (e.g. 2B, 2C)

Here, we employed annexin V (AnnV) and propidium iodide (PI) staining to evaluate apoptosis and cell death in Ly6Chi and Ly6Clo blood monocytes from control and RBPJ deficient mice. The results showed no significant difference in the levels of apoptosis and cell death between the two groups (see Author response image 2 below). The statistical data for Ki-67 expression obtained from multiple experiments, and the expression of Ki-67 showed no significant difference between the control and RBP-J deficient mice (see the figure B below). In Figure 2C, each dot represents 2-3 mice, and there were no differences observed between control and RBP-J deficient mice at multiple time points during the repeated measurements.

1. Sometimes strong statements were based on the lack of statistical significance, when more n number could have changed the interpretation (e.g. 2G, 3E)

We have derived the corresponding conclusions based on the observed experimental results.

1. There is incomplete analysis (e.g. Network analysis) and interpretation of RNAsequencing results (figure 4), the difference between the genotypes in both monocyte subsets would provide a more complete picture and potentially reveal mechanisms

We thank the reviewer for bringing out this point. We agreed that a more comprehensive analysis, including a comparison between the genotypes in both monocyte subsets, would provide a deeper understanding and potentially uncover underlying mechanisms. Having observed alterations in blood Ly6Clo monocytes in RBP-J deficient mice, our primary focus had been on analyzing the differentially expressed genes within this subset of monocytes to gain further insights into its specific characteristics and behavior. We also uploaded sequencing data sets in the Genome Expression Omnibus with assigned accession numbers GSE208772 to facilitate interested researchers in accessing and downloading the data.

1. The experiments in Figures 5 and 7 are missing a control (Lyz2cre/cre Ccr2RFP/RFP or the Rbpj+/+ versions) and may have been misinterpreted. For example if the control (RBP-J WT, CCR2 KO) was used then it would almost certainly show falling Ly6C low numbers compared to RBP-J WT CCR2 WT, but RBP-J KO CCR2 KO would still have more Ly6c low monocytes than RBP-J WT, CCR2 KO - meaning that the RBP-J function is independent of CCR2. I.e. Ly6c low numbers are mostly dependent on CCR2 but this is irrespective of RBP-J.

The diminished Ly6Clo monocytes in Rbpjfl/flLyz2cre/creCcr2RFP/RFP (DKO) mice can be divided into two distinct subpopulations: one portion originates from Ly6Chi monocytes, while the other comprises Ly6Clo monocytes characterized by heightened CCR2 expression. The Ly6Clo monocytes that remain in DKO mice exhibit CCR2 expression levels within the normal range when compared to Lyz2cre/cre mice, but lower levels compared to RBP-J deficient mice (Figure 5A). These findings suggest that RBP-J exerts regulatory influence over Ly6Clo monocytes, at least in part, through CCR2.

1. Figure 6 was difficult to interpret because of the lack of shown gating strategy. This reviewer assumes that alveolar macrophages were gated out of analysis

The gating strategy of lung interstitial macrophage in the manuscript Figure 6 was consistent with the published work (Schyns et al, cited in the manuscript). We also measured alveolar macrophages (AM) from control and RBP-J deficient mice bronchoalveolar lavage fluid. At the resting state, RBP-J deficient mice exhibited normal AM frequency and number (see Author response image 3 below).

1. The statements around Figure 7 are not completely supported by the evidence, i) a significant proportion of CD16.2+ cells were CCR2 independent and therefore potentially not all recently derived from monocytes, and ii) there is nothing to suggest that the source was not Ly6C high monocytes that differentiated - the manuscript in general seems to miss the point that the source of the Ly6C low cells is almost certainly the Ly6C high monocytes - which further emphasises the importance of both cells in the sequencing analysis

Schyns et al and Sabatel at al showed that the numbers of IM and CD16.2+ were similar in Ccr2 sufficient and Ccr2-/- mice, demonstrating that CD16.2+ cells were Ccr2 independent. The number of CD16.2+ cells was significantly reduced in Rbpjfl/flLyz2cre/creCcr2RFP/RFP mice as compared to Rbpjfl/flLyz2cre/cre mice, in line with decreased number of lung Ly6Clo monocytes and blood Ly6Clo monocytes, showing that CD16.2+ cells depended on Ccr2 for their presence in Rbpjfl/flLyz2cre/cre mice.

1. The authors did not refer to or cite a similar 2020 study that also investigated myeloid deletion of Rbpj (Qin et al. 2020 - https://doi.org/10.1096/fj.201903086RR). Qin et al identified that Ly6Clo alveolar macrophages were decreased in this model - it is intriguing to synthesise these two studies and hypothesise that the ly6c low monocytes steal the lung niche, but this was not discussed

We thank the reviewer for bringing this study to our attention. According to their findings, myeloid-specific RBP-J deficiency resulted in a decrease in Ly6CloCD11bhi alveolar macrophages but an increase in Ly6CloCD11blo alveolar macrophages after bleomycin treatment, while the total number of alveolar macrophages showed no significant difference. These results suggest that RBP-J may play a role in regulating the balance between these specific alveolar macrophage subsets in response to bleomycin-induced injury, without affecting the overall population of alveolar macrophages. This may be different from what we observe in interstitial macrophages under resting conditions.

Reviewer #3

1. It is curious that the authors do not see the increase in circulating monocytes reflected in the spleen however, the n-number is 2. Increasing the n-number would enable the author to understand the data which is not interpretable at the moment. There are multiple other places in which a low n-number makes it hard to fully understand the biology (eg Figure 2C&E)

Although we only counted the number of splenic monocyte subsets in two mice, the proportion of splenic monocyte subsets was calculated based on additional quantity of mice in our study.

1. Given that Ly6Clow monocytes are thought to be longer lived than Ly6C+ and there is still considerable labelling of Ly6Clow monocytes at the end of the 96 hours analysed in the EdU experiment, it is not possible to determine from the data here whether RBPJ deficiency increases life span. Could it be that differences in %EdU+ cells would only be seen at later time points? If the timeline was extended, could it be that differences in %EdU+ become apparent

Based on the latex bead experiment, we observed that the presence of latex+ Ly6Clo monocytes at 7 days in control and RBP-J deficient mice did not differ, indicating that the lifespan of Ly6Clo monocytes did not increase.

1. Similarly for the latex bead experiment. Given that there is only n=2 at the first time point and only ~30% of Ly6Clow monocytes are Latex+, it is very hard to conclusively claim that RBP-J does not influence monocyte survival or proliferation. An interesting experiment to assess whether RBP-J is increasing monocyte survival could be an adoptive transfer model in which Ly6Clow monocytes are injected into a congenic mouse and tracked over time.

In RBP-J deficient mice, there was an increase in the proportion of Ly6Clo monocytes. We hypothesized that this lower proportion of latex+ cells might make it easier to observe differences, but clearly, in our experiment, no differences were observed between control and RBP-J deficient mice.

1. RNA-seq: Ccr2 and Itgax are not the top hits. The authors do not investigate the top hits which may provide very interesting insight into how RBP-J influences monocyte biology.

We thank the reviewer for raising these points. We also analyzed some top changed genes. The top two gene in the downregulated gene list are Hes1 and Nrarp, which are regulated by the Notch pathway (Krebs et al 2001 and Radtke et al 2010). We tested blood monocytes, but the population of monocyte subsets displayed no differences between Hes1fl/flRbp-jfl/flLyz2cre/cre and Rbp-jfl/flLyz2cre/cre mice (data not shown). As shown in Figure 2- figure supplement 1A, expression of Nr4a1 showed no significant differences between control and RBP-J deficient mice. The top gene in the upregulated gene list is Erdr1, which has been reported to play a role in cellular survival (Soto et al 2017), while blood monocyte subsets in RBP-J deficient mice displayed normal survival.

1. The PCA plot in figure 4C- it would be interesting to see where all the biological replicates fall.

We agree with the reviewer’s assessment that observing the positions of all biological replicates on the PCA plot may indeed yield valuable insights. However, it is worth noting that the upregulated and downregulated genes also offer suggestive hints.

1. Based on CCR2 expression and CD11c expression, monocytes from RBP-J deficient mice look more like Ly6C+ monocytes - could it be that RBP-J is increasing conversion from Ly6C+ monocytes to Ly6Clow? Or could it be that Ly6Clow monocytes are heterogeneous and RBP-J is increasing survival or conversion of one subtype of Ly6Clow monocytes but looking at all Ly6Clow monocytes together is masking this?

Ly6Clo monocyte can be subdivided into different subpopulations depending on surface makers, such as CD43, MHC-II, CD11c and CCR2 (Jakubzick et al 2013 and Ginhoux et al. 2014). Carlin et al founded that a subset of blood Ly6Clow cells was independent of both Ccr2 and Nr4a1. As said by the reviewer, Ly6Clo monocytes are heterogeneous. Therefore, there is a possibility of altered survival in a certain group of Ly6Clo monocytes.

1. The data presented here suggest that lung CD16.2+ interstitial macrophages are derived from Ly6Clow monocytes which are increased via CCR2. Although the data are suggestive, they are not conclusive, lineage tracing and CCR2 blockade or better, conditional CCR2 deficiency would help to strengthen the claim.

Schyns et al showed that the number of CD16.2+ was similar in Ccr2 sufficient and Ccr2-/- mice, demonstrating that CD16.2+ cells were Ccr2 independent. While number of CD16.2+ cells was significantly reduced in Rbpjfl/flLyz2cre/creCcr2RFP/RFP mice as compared to Rbpjfl/flLyz2cre/cre mice, in line with decreased number of lung Ly6Clo monocytes and blood Ly6Clo monocytes. Moreover, the turnover of lung Ly6Chi and Ly6Clo monocytes was normal. These results implicated that CD16.2+ cells depended on Ccr2 for their presence in Rbpjfl/flLyz2cre/cre mice.

1. The figures could do with more headings/ more detailed legends to help the reader, for example including what is BM, what is blood, what is spleen. Figure 2E needs the days labelled on or above the histograms.

We thank the reviewer for raising this important point. We have now added additional detailed legends to the figure.

1. Gating strategies should be included to help the reader understand which cells you are looking at, especially for Figure 6&7.

The gating strategy for Figures 6 and 7 followed the method reported in the literature, which included the identification of alveolar macrophages. Additionally, we labeled the markers for cell populations in the figure.

Version 2

Abstract

Abstract

Notch-RBP-J signaling plays an essential role in maintenance of myeloid homeostasis. However, its role in monocyte cell fate decisions is not fully understood. Here we showed that conditional deletion of transcription factor RBP-J in myeloid cells resulted in marked accumulation of blood Ly6Clo monocytes that highly expressed chemokine receptor CCR2. Bone marrow transplantation and parabiosis experiments revealed a cell intrinsic requirement of RBP-J for controlling blood Ly6CloCCR2hi monocytes. RBP-J-deficient Ly6Clo monocytes exhibited enhanced capacity competing with wildtype counterparts in blood circulation. In accordance with alterations of circulating monocytes, RBP-J deficiency led to markedly increased population of lung tissues with Ly6Clo monocytes and CD16.2+ interstitial macrophages. Furthermore, RBP-J deficiency-associated phenotypes could be genetically corrected by further deleting Ccr2 in myeloid cells. These results demonstrate that RBP-J functions as a crucial regulator of blood Ly6Clo monocytes and thus derived lung-resident myeloid populations, at least in part through regulation of CCR2.

Introduction

Monocytes are integral components of the mononuclear phagocyte system (MPS) that develop from monocyte precursors in the bone marrow(Guilliams et al., 2018). Several functionally and phenotypically distinct subsets of blood monocytes have been defined on the basis of expression of surface markers(Cros et al., 2010; Geissmann et al., 2003; Passlick et al., 1989; Weber et al., 2000). In mice, Ly6Chi monocytes (also called inflammatory or classical monocytes) characterized by Ly6ChiCCR2hiCX3CR1lo exhibit a short half-life, and are recruited to tissue and differentiate into macrophages and dendritic cells(Ginhoux and Jung, 2014). Ly6Chi monocytes are analogous to CD14+ human monocytes based on gene expression profiling(Ingersoll et al., 2010). A second subtype of monocytes is called Ly6Clo monocytes (also termed patrolling monocytes or non-classical monocytes), which express high level of CX3CR1 and low level of CCR2, equivalent to CD14loCD16+ human monocytes. Ly6Clo monocytes are regarded as blood-resident macrophages that patrol blood vessels and scavenge microparticles attached to the endothelium under physiological conditions, and exhibit a long half-life in the steady-state(Auffray et al., 2007; Carlin et al., 2013; Yona et al., 2013). Ly6Clo monocytes may exert a protective effect by suppressing atherosclerosis in mice, while high levels of non-classical monocytes have been associated with more advanced vascular dysfunction and oxidative stress in patients with coronary artery disease(Hamers et al., 2012; Hanna et al., 2012; Urbanski et al., 2017). In the inflammatory settings, Ly6Clo monocytes often show anti-inflammatory properties, yet also exhibit a proinflammatory role under certain circumstances(Amano et al., 2005; Brunet et al., 2016; Carlin et al., 2013; Cros et al., 2010; Misharin et al., 2014; Nahrendorf et al., 2007; Puchner et al., 2018). In addition, Ly6Clo monocytes have demonstrated protective functions in tumorigenesis, such as engulfing tumor materials to prevent cancer metastasis, activating and recruiting NK cells to the lungs(Hanna et al., 2015; Thomas et al., 2016). While the functions of Ly6Clo monocytes are complex and sometimes contradictory depending on the animal models and experimental approaches, it is clear that they play a crucial role in both health and disease.

The generation and maintenance of Ly6Clo monocytes are regulated by several transcription factors, including Nr4a1 and CEBPβ(Hanna et al., 2011; Tamura et al., 2017). The colony-stimulating factor 1 receptor (CSF1R) signaling pathway is important for the generation and survival of Ly6Clo monocytes, and neutralizing antibodies against CSF1R can reduce their numbers(MacDonald et al., 2010). Interestingly, recent studies have suggested that different subsets of monocytes may be supported by distinct cellular sources of CSF-1 within bone marrow niches, in which targeted deletion of Csf1 from sinusoidal endothelial cells selectively reduces Ly6Clo monocytes but not Ly6Chi monocytes(Emoto et al., 2022). LAIR1 has been shown to be activated by stromal protein Colec12, and LAIR1 deficiency leads to aberrant proliferation and apoptosis in bone marrow non-classical monocytes(Keerthivasan et al., 2021). These findings further underscore the complexity of the mechanisms that regulate Ly6Clo monocytes, and suggest that multiple factors are involved in this process.

Recombination signal-binding protein for immunoglobulin kappa J region (RBP-J; also named CSL in humans) is commonly known as the master nuclear mediator of canonical Notch signaling(Radtke et al., 2013). In the absence of Notch intracellular domain, RBP-J associates with co-repressor proteins to repress transcription of downstream target genes. In the immune system, the best studied functions of Notch signaling are its roles in regulating the development of lymphocytes, such as T cells and marginal zone B cells(Radtke et al., 2013). Notch signaling also has been reported to regulate the differentiation and function of myeloid cells including granulocyte/monocyte progenitors, osteoclasts and dendritic cells (Caton et al., 2007; Klinakis et al., 2011; Lewis et al., 2011; Zhao et al., 2012). In osteoclasts, RBP-J represses osteoclastogenesis, particularly under inflammatory conditions(Zhao et al., 2012). In dendritic cells, Notch-RBP-J signaling controls the maintenance of CD8- DC(Caton et al., 2007). In addition, Notch2-RBP-J pathway is essential for development of CD8-ESAM+ DC in the spleen and CD103+CD11b+ DC in the lamina propria of the intestine(Lewis et al., 2011). However, the role of Notch-RBP-J signaling pathway in regulating monocyte subsets remains elusive. Here, we demonstrated that RBP-J controlled homeostasis of blood Ly6Clo monocytes in a cell intrinsic manner. RBP-J deficiency led to a drastic increase in blood Ly6Clo monocytes, lung Ly6Clo monocytes and CD16.2+ IM. Our results suggest that RBP-J plays a critical role in regulating the population and characteristics of Ly6Clo monocytes in the blood and lung.

Results

RBP-J is essential for maintenance of blood Ly6Clo monocytes

To investigate the role of RBP-J in monocyte subsets, we utilized mice with RBP-J specific deletion in the myeloid cells (Rbpjfl/flLyz2cre/cre mice). Efficient deletion of Rbpj in blood monocytes was confirmed by quantitative real-time PCR (qPCR) (Figure 1 – figure supplement 1A). Flow cytometric analysis revealed that RBP-J deficient mice had a significant increase in the proportion of Ly6Clo monocytes, but not Ly6Chi monocytes, in blood, compared to age-matched control mice with the genotype Rbpj+/+Lyz2cre/cre (Figure 1A). In contrast to circulating monocytes, RBP-J deficient mice exhibited minimal alterations in the percentages of monocyte subsets in bone marrow (BM) and spleen (Figure 1B and Figure 1 – figure supplement 1C). Additionally, RBP-J deficiency in myeloid cells did not appear to have an effect on neutrophils (Figure 1– figure supplement 1B). Therefore, among circulating myeloid populations, RBP-J selectively controlled the subset of Ly6Clo monocytes.

RBP-J deficient mice display more Ly6Clo blood monocytes.

(A) Blood Ly6Chi and Ly6Clo monocytes in Rbpj+/+Lyz2cre/cre control and Rbpjfl/flLyz2cre/cre mice were determined by flow cytometry analyses (FACS). Representative FACS plots (left) and cumulative data of cell ratio (right) are shown.

(B and C) Representative FACS plots and cumulative data quantitating percentages of BM monocyte subsets (B) and myeloid progenitor cells (C) (CD45+CD11b+Ly6G- CD115+ Ly6Chi monocyte; CD45+CD11b+Ly6G-CD115+Ly6Clo monocyte; MDP, CD45+Lin-CD117+CD115+CD135+ Ly6C-; cMoP, CD45+Lin-CD11b-CD117+CD115+ CD135-Ly6C+). Lin: CD3, B220, Ter119, and Gr-1, CD11b.

(D) Experimental outline for panel (E).

(E) Cumulative data quantitating percentages of sinusoidal monocytes (CD45+) within total BM Ly6Clo monocytes.

Data are pooled from at least 2 independent experiments; n ≥ 4 in each group. Data are shown as mean ± SEM; n.s., not significant; **, P < 0.01 (two-tailed Student’s unpaired t test). Each symbol represents an individual mouse.

BM progenitors that give rise to circulating monocytes are monocyte-dendritic cell progenitors (MDPs) and common monocyte progenitors (cMoPs)(Hanna et al., 2011; Hettinger et al., 2013; Liu et al., 2019; Varol et al., 2007). Next, we analyzed BM progenitors, and found that the percentages of MDPs and cMoPs were equivalent in control and RBP-J deficient mice (Figure 1C). These results in conjunction with the observations of normal BM monocyte populations suggest that alterations of peripheral blood Ly6Clo monocytes may not originate from BM. Next, we examined whether RBP– J may influence the egress of Ly6Clo monocytes from BM, and measured BM exit rate of Ly6Clo monocytes using in vivo labeling of sinusoidal cells as previously described (Figure 1D)(Debien et al., 2013). The results showed that RBP-J deficiency did not affect egress of Ly6Clo monocytes from BM (Figure 1E). Taken together, these data revealed RBP-J as a critical regulator controlling homeostasis of peripheral blood Ly6Clo monocytes.

RBP-J deficiency does not affect Ly6Chi monocyte conversion or Ly6Clo monocyte survival and proliferation

Next, we aimed to determine whether the increase of blood Ly6Clo monocytes in RBP– J deficient mice was due to decreased cell death or enhanced proliferation. We first stained monocytes with Annexin V and 7-AAD to identify apoptotic cells and observed comparable percentages of apoptotic blood Ly6Clo monocytes in control and RBP-J deficient mice (Figure 2A). Given the crucial role of Nr4a1 in the survival of Ly6Clo monocytes(Hanna et al., 2011), we detected the expression of Nr4a1, which was similar in Ly6Clo monocytes from control and RBP-J deficient mice (Figure 2 – figure supplement 1A). We then assessed the proliferative capacity by analyzing Ki-67 expression as well as in vivo EdU incorporation. Ki-67 levels in blood monocytes displayed no differences between control and RBP-J deficient mice (Figure 2B). The percentage of EdU+ monocytes did not significantly differ between control and RBP-J deficient mice at the indicated time points, implying that the turnover of Ly6Clo monocytes was normal in RBP-J deficient mice (Figure 2C). Fluorescent latex beads as particulate tracers could be phagocytosed by monocytes after intravenous injection, and stably label Ly6Clo monocytes(Tacke et al., 2006). Thus, we intravenously injected fluorescent latex beads into control and RBP-J deficient mice to track circulating monocytes (Figure 2D). By day 7 post injection, only Ly6Clo monocytes were latex+ as previously reported(Tacke et al., 2006), whereas control and RBP-J deficient mice groups presented a similar frequency of latex+ monocytes (Figure 2E). Together, these data indicated that RBP-J did not influence monocyte survival and proliferation.

Monocyte subsets in RBP-J deficient mice display normal cell death.

(A) Representative FACS plots of monocyte subsets in blood stained with 7-AAD and Annexin V.

(B) FACS analysis of Ki-67 expression in Rbpj+/+Lyz2cre/cre control and Rbpjfl/flLyz2cre/cre blood monocyte subsets. Black lines represent control mice and red lines represent RBP-J deficient mice.

(C) Analysis of time course of EdU incorporation of monocyte subsets in BM and blood after a single 1 mg EdU pulsing. The percentages of EdU+ cells among the indicated monocyte subsets are shown.

(D) Experimental outline for panel (E).

(E) Analysis of time course of latex beads incorporation of monocyte subsets in blood after latex beads injection. The percentages of latex+ cells among the indicated monocyte subsets are shown.

(F) Cartoon depicting the adoptive transfer. BM GFP+Ly6Chi monocytes were sorted from Lyz2cre/creCx3cr1gfp/+or Rbpjfl/flLyz2cre/creCx3cr1gfp/+mice and transferred into Ccr2RFP/RFP recipient mice. Sixty hours after transfer, cell fate was analyzed.

(G) Representative FACS plots are shown in the left panel, and the frequencies of GFP+Ly6Clo monocytes within total Ly6Clo monocytes are shown in the right panel.

Data are pooled from 2 independent experiments (G); n ≥ 2 in each group (C, E and G). Data are shown as mean ± SEM; n.s., not significant; (two-tailed Student’s unpaired t test). Each symbol represents an individual mouse (E and G).

Previous studies have shown that Ly6Clo monocytes are observed in recipient mice following the adoptive transfer of Ly6Chi monocytes, indicating that Ly6Chi monocytes can convert into Ly6Clo monocytes(Varol et al., 2007; Yona et al., 2013). We next wished to evaluate whether conversion of Ly6Chi monocyte was regulated by RBP-J by isolating bone marrow GFP+Ly6Chi monocytes from Lyz2cre/creCx3cr1gfp/+ control or Rbpjfl/flLyz2cre/creCx3cr1gfp/+ mice and adoptively transferring them into Ccr2RFP/RFP recipients (Figure 2F). Sixty hours after transfer, a subset of cells from donors were converted into Ly6Clo monocytes (Figure 2 – figure supplement 1B), and equal percentages of Ly6Clo monocytes were derived from control and Rbpjfl/flLyz2cre/creCx3cr1gfp/+donors (Figure 2G). Thus, the conversion of Ly6Chi monocyte into Ly6Clo monocyte was not affected by RBP-J deficiency.

RBP-J regulates blood Ly6Clo monocytes in a cell intrinsic manner

We next wondered whether the increase of Ly6Clo monocytes in RBP-J deficient mice was bone marrow-derived and cell intrinsic. We performed bone marrow transplantation by engrafting lethally irradiated mice with a 1:4 mixture of Rbpj+/+Lyz2cre/cre control and Rbpjfl/flLyz2cre/cre BM cells (CD45.2) and Cx3cr1gfp/+ BM cells (CD45.1) (Figure 3A). Eight weeks after transplantation, we analyzed the frequencies of CD45.2+ donor cells in the bone marrow and blood of recipient mice and found that the frequencies of CD45.2+ cells within total cells were similar to the mixture ratio of bone marrow cells for Ly6Chi monocytes and neutrophils in both BM and blood (Figure 3B and Figure 3 – figure supplement 1A and B). Specifically, more blood Ly6Clomonocytes were derived from RBP-J deficient donors than control donors (Figure 3B), reflecting that the contribution of RBP-J deficient cells in blood Ly6Clo cells was significantly higher than that of control cells. These results implied that RBP-J deficient BM cells were highly efficient in generating blood Ly6Clo monocytes.

The role of RBP-J in blood Ly6Clo monocytes is cell intrinsic.

(A) Cartoon depicting the bone marrow transplantation. BM recipient 6-week-old male C57BL/6 mice (CD45.2) were lethally irradiated. BM cells from donor mice (Rbpj+/+Lyz2cre/cre or Rbpjfl/flLyz2cre/cre and CD45.1, Cx3cr1gfp/+) were collected and transferred into recipient mice. Mice were used after 8 weeks of bone marrow reconstitution.

(B) Representative FACS plots (left) and cumulative data (right) quantitating the frequency of Rbpj+/+Lyz2cre/cre and Rbpjfl/flLyz2cre/cre donor cells among Ly6Chi and Ly6Clo monocytes in the blood of recipient mice.

(C) Cartoon depicting the generation of Rbpj+/+Lyz2cre/cre control or Rbpjfl/flLyz2cre/cre and Cx3cr1gfp/+ parabiotic pairs.

(D) Representative FACS plots (left) and cumulative data (right) quantitating percentages of monocyte subsets derived from control or RBP-J deficient mice in control or RBP-J deficient mice.

(E) Representative FACS plots (left) and cumulative data (right) quantitating percentages of monocyte subsets derived from control or RBP-J deficient mice in Cx3cr1gfp/+ mice.

Data are pooled from at least 2 independent experiments; n ≥ 6 in each group. Data are shown as mean ± SEM; n.s., not significant; ***, P < 0.001; ****, P < 0.0001 (two– tailed Student’s unpaired t test). Each symbol represents an individual mouse. SSC-A, side scatter area.

Given that RBP-J deficient mice had more Ly6Clo monocytes in blood than did control mice, we performed parabiosis experiments, a surgical union of two organisms, which allowed parabiotic mice to share their blood circulation. The contribution of circulating cells from one animal in another can be estimated by measuring the percentage of blood cells that originated from each animal(Liu et al., 2007). We joined a CD45.1, Cx3cr1gfp/+ mouse with an age and sex-matched Rbpj+/+Lyz2cre/cre control or Rbpjfl/flLyz2cre/cre mouse (CD45.2) (Figure 3C). Four weeks after the procedure, about 50% of B cells and T cells in parabiotic mice displayed efficient exchange of their circulation (Figure 3 – figure supplement 1C). As expected, RBP-J deficient mice exhibited higher percentages of Ly6Clo monocytes than control animals (Figure 3D). Intriguingly, in the parabiotic Cx3cr1gfp/+mice, RBP-J deficient cells still constituted significantly higher proportion of circulating Ly6Clo, but not Ly6Chi monocytes, than RBP-J sufficient counterparts as indicated by the percentages of the GFP-negative population (Figure 3E). These results implicated that the enhanced ability of Ly6Clo monocytes to circulate in the peripheral blood as a result of RBP-J deficiency was cell intrinsic.

RBP-J regulates phenotypical marker genes in blood Ly6Clo monocytes

To further study the consequences of RBP-J loss of function in blood Ly6Clo monocytes, we performed gene expression profiling by RNA-seq, which revealed that Ly6Clo monocytes in RBP-J deficient mice exhibited low expression of Itgax but high expression of Ccr2, in comparison to those in control mice (Figure 4A). These differential expression patterns of phenotypic marker genes were also confirmed by qPCR and flow cytometry analysis (Figure 4B). However, Ly6Chi blood monocytes and BM monocytes displayed normal levels of CD11c and CCR2 (Figure 4B and Figure 4 – figure supplement 1A). Moreover, principal component analysis showed distinct expression pattern of monocytes between control and RBP-J deficient mice (Figure 4C). To determine whether the regulation of phenotypic markers by RBP-J in blood Ly6Clo monocytes was cell-intrinsic, we performed BM transplantation as described above (Figure 3A). The findings indicated that Ly6Clo monocytes derived from RBP-J deficient BM cells expressed high levels of CCR2 but low levels of CD11c, in comparison to those derived from control BM cells, whereas no significant changes were observed in blood Ly6Chi and BM monocytes, as expected (Figure 4D and Figure 4 – figure supplement 1B). These results collectively suggested that RBP-J regulated the expression of CCR2/CD11c in a cell-intrinsic manner.

Phenotypic markers in Ly6Clo monocytes are changed in RBP-J deficient mice.

(A) Heatmap of RNA-seq dataset showing the top 20 downregulated and upregulated genes in blood Ly6Clo monocytes from Rbpjfl/flLyz2cre/cre versus Rbpj+/+Lyz2cre/cre control mice. Blue and red font indicates downregulated and upregulated genes in RBP-J deficient Ly6Clo monocytes respectively.

(B) Representative FACS plots, cumulative MFI, and qPCR analysis of CCR2/CD11c expression in control and RBP-J deficient blood monocyte subsets. Shaded curves represent isotype control, black lines represent control mice and red lines represent RBP-J deficient mice.

(C) PCA of indicated cell types.

(D) Representative FACS plots and cumulative MFI of CCR2/CD11c expression in blood monocyte subsets derived from control or RBP-J deficient mice. Black lines represent control mice, and red lines represent RBP-J deficient mice.

Data are pooled from 2 independent experiments (B and D); n ≥ 6 in each group. Data are shown as mean ± SEM; n.s., not significant; **, P < 0.01; ****, P < 0.0001 (two– tailed Student’s unpaired t test). Each symbol represents an individual mouse.

RBP-J-mediated control of blood Ly6Clo monocytes is CCR2 dependent

Given that RBP-J deficiency led to enhanced CCR2 expression in Ly6Clo monocytes and that CCR2 is essential for monocyte functionality under various inflammatory and non-inflammatory conditions(Shi and Pamer, 2011), we wished to genetically investigate the role of CCR2 in the RBP-J deficient background. We generated RBP– J/CCR2 double deficient (DKO) mice with the genotype Rbpjfl/flLyz2cre/creCcr2RFP/RFPwith Lyz2cre/creCcr2RFP/+ and Rbpjfl/flLyz2cre/creCcr2RFP/+mice serving as controls. As expected, the expression of CCR2 in Ly6Clo monocytes was reduced in DKO mice compared to RBP-J deficient mice (Figure 5A), confirming the successful deletion of the Ccr2 gene. DKO mice showed a lower percentage of both Ly6Chi and Ly6Clo monocytes than RBP-J deficient mice (Figure 5B and C). Notably, the percentage of Ly6Clo monocytes in DKO mice was comparable to that observed in the control mice (Figure 5B and C), implicating that deletion of CCR2 corrected RBP-J deficiency– associated phenotype of increased Ly6Clo monocytes. Whereas, both RBP-J deficient and DKO mice exhibited lower expression levels of CD11c in their Ly6Clo monocyte than control mice, suggesting that the regulation of CD11c expression was independent of CCR2 (Figure 5D). These results suggested that RBP-J regulated Ly6Clo monocytes, at least in part, through CCR2.

Blood Ly6Clo monocytes are decreased in DKO mice.

(A) Representative FACS plots and cumulative MFI of CCR2 expression in Lyz2cre/creCcr2RFP/+ control, Rbpjfl/flLyz2cre/creCcr2RFP/+ and Rbpjfl/flLyz2cre/creCcr2RFP/RFP (DKO) blood Ly6Chi and Ly6Clo monocytes are shown.

(B and C) Blood monocyte subsets in control, RBP-J deficient and DKO mice were determined by FACS. Representative FACS plots (B) and cumulative data of cell ratio

(C) are shown.

(D) Representative FACS plots and cumulative MFI of CD11c expression in control, RBP-J deficient and DKO blood Ly6Chi and Ly6Clo monocytes are shown.

Data are pooled from at least 2 independent experiments; n ≥ 5 in each group. Data are shown as mean ± SEM; n.s., not significant; **, P < 0.01; ***, P < 0.001; ****, P < 0.0001 (two-tailed Student’s unpaired t test). Each symbol represents an individual mouse.

Lung Ly6Clo monocytes are accumulated in RBP-J deficient mice

In mice, alveolar macrophages (AM) are maintained by local self-renewal, and arise from fetal liver derived precursors, whereas interstitial macrophages (IM) are probably originated from monocytes (Guilliams et al., 2013; Hashimoto et al., 2013; Sabatel et al., 2017). Schyns et al identified 3 subpopulations of IM according to expression of CD206 and CD16.2 and Ly6Clo monocytes are proposed to give rise to CD64+CD16.2+ IM(Schyns et al., 2019). We next compared the populations of lung monocytes and IM between control and RBP-J deficient mice. The conditional deletion of RBP-J resulted in a significant increase in the absolute and relative numbers of Ly6Clo monocytes and CD16.2+ IM, while the numbers of Ly6Chi monocytes, CD206- IM, and CD206+ IM remained unchanged (Figure 6A and B). To confirm these findings, lung sections from Lyz2cre/creCx3cr1gfp/+ control and Rbpjfl/flLyz2cre/creCx3cr1gfp/+ mice were stained with an anti-GFP antibody, which revealed a significant increase in the number of GFP+ cells in RBP-J deficient mice (Figure 6C). In addition, we evaluated the proliferative capacity of Ly6Clo monocytes and CD16.2+ IM, but did not observe enhanced EdU incorporation in Ly6Clo monocytes and CD16.2+ IM from RBP-J deficient mice (Figure 6 – figure supplement 1A and B). These findings suggested that the augmented population of these cells might not be resulted from increased in situ proliferation. Previous reports have indicated that LPS can increase the population of IM in a CCR2-dependent manner(Sabatel et al., 2017). We thus challenged mice with LPS, and analyzed monocytes and IM at day 0 and day 4. LPS exposure induced an increase in numbers of Ly6Clo monocytes, CD16.2+ IM and CD206- IM compared to baseline both in control and RBP-J deficient mice (Figure 6D and E). There was a trend toward higher Ly6Chi monocyte after LPS treatment, although this observation was not statistically significant. Of note, RBP-J deficient mice exhibited robustly elevated numbers of Ly6Clo monocytes and CD16.2+ IM compared with control mice after LPS treatment (Figure 6D and E), suggesting that RBP-J deficient Ly6Clo monocytes were recruited to inflamed tissue in large numbers and differentiated into CD16.2+ IM.

RBP-J deficient mice exhibit more lung Ly6Clo monocytes and CD16.2+ IM.

(A and B) Indicated populations in the lungs of Rbpj+/+Lyz2cre/cre and Rbpjfl/flLyz2cre/cre mice were determined by FACS. Representative FACS plots (A) and cumulative data of cell ratio and absolute numbers (B) are shown.

(C) Immunofluorescence staining for GFP+ cells in the lungs from Lyz2cre/creCx3cr1gfp/+and Rbpjfl/flLyz2cre/creCx3cr1gfp/+mice (CX3CR1 [green]; DAPI [blue]). Scale bars represent 50 µm.

(D and E) Rbpj+/+Lyz2cre/cre and Rbpjfl/flLyz2cre/cre mice were instilled intranasally with PBS or PBS containing LPS, and lungs were harvested at the indicated time points. Representative FACS plots (D) and cumulative data of cell ratio and absolute numbers

(E) are shown.

Data are pooled from at least 2 independent experiments; n ≥ 4 in each group. Data are shown as mean ± SEM; n.s., not significant; *, P < 0.05; **, P < 0.01; ****, P < 0.0001 (two-tailed Student’s unpaired t test). Each symbol represents an individual mouse.

To investigate whether the elevated levels of lung monocytes and CD16.2+ IM in RBP-J deficient mice were derived from blood Ly6Clo monocytes, we examined monocytes and IM in Lyz2cre/creCcr2RFP/+control, Lyz2cre/creCcr2RFP/RFP, Rbpjfl/flLyz2cre/creCcr2RFP/+, and Rbpjfl/flLyz2cre/creCcr2RFP/RFP(DKO) mice. The deletion of CCR2 led to a decrease in Ly6Chi monocytes, and the Ly6Clo monocytes in DKO mice were reduced to a level similar to that of control mice (Figure 7A and B). While the DKO mice had more CD16.2+ IM compared to control mice, these cells were severely reduced compared to RBP-J deficient mice (Figure 7A and C). The above data supported the notion that increased lung Ly6Clo monocytes and CD16.2+ IM in RBP-J deficient mice were derived from Ly6Clo blood monocytes. In summary, these results suggested that in RBP-J deficient mice, recruitment of blood Ly6Clo monocytes to the lung was markedly facilitated by increased cell number as well as heightened expression of CCR2, leading to the increase of lung Ly6Clo monocytes and CD16.2+ IM.

DKO mice lack lung Ly6Clo monocytes and CD16.2+ IM.

(A) Representative FACS plots of lung monocyte and IM subsets in Lyz2cre/creCcr2RFP/+, Rbpjfl/flLyz2cre/creCcr2RFP/+, Lyz2cre/creCcr2RFP/RFP and Rbpjfl/flLyz2cre/creCcr2RFP/RFP mice.

(B and C) Cumulative data of cell ratio and absolute numbers of monocyte (B) and IM

(C) subsets.

Data are pooled from 2 independent experiments; n ≥ 4 in each group. Data are shown as mean ± SEM; n.s., not significant; *, P < 0.05; **, P < 0.01; ***, P < 0.001; ****, P < 0.0001 (two-tailed Student’s unpaired t test). Each symbol represents an individual mouse.

Discussion

In this study, we identified RBP-J as a pivotal factor in regulating blood Ly6Clo monocyte cell fate. Our results showed that mice with conditional deletion of RBP-J in myeloid cells exhibited a robust increase in blood Ly6Clo monocytes, and subsequently accumulated lung Ly6Clo monocytes and CD16.2+ IM under steady-state conditions. Bone marrow transplantation experiments in which RBP-J deficient cells were transplanted into recipient mice as well as the parabiosis experiment showed similar increase in blood Ly6Clo monocytes, demonstrating that RBP-J was an intrinsic factor required for the regulation of Ly6Clo monocytes. Further analysis revealed that RBP-J regulated the expression of CCR2 and CD11c on the surface of Ly6Clo monocytes. Moreover, the phenotype of elevated blood Ly6Clo monocytes and their progeny was further ameliorated by deleting Ccr2 in a RBP-J deficient background.

Notch-RBP-J signaling has been shown to play a role in regulating functional polarization and activation of macrophages, and regulated formation of Kupffer cells and macrophage differentiation from Ly6Chi monocytes in ischemia(Foldi et al., 2016; Hu et al., 2008; Kang et al., 2020; Krishnasamy et al., 2017; Sakai et al., 2019; Wang et al., 2010; Xu et al., 2012). At steady-state, interaction of DLL1 with Notch 2 regulates conversion of Ly6Chi monocytes into Ly6Clo monocytes in special niches of the BM and spleen(Gamrekelashvili et al., 2016). Under inflammatory conditions, Notch2 and TLR7 pathways independently and synergistically promote conversion of Ly6Chi monocytes into Ly6Clo monocytes(Gamrekelashvili et al., 2020). However, in our study, the percentage of blood Ly6Clo monocytes increased in RBP-J deficient mice. These results somewhat differ from what was observed in mice with conditional deletion of Notch2(Gamrekelashvili et al., 2016). Actually, 4 members of the Notch family have been identified in mammals(Radtke et al., 2013). Thus, Notch receptors seem to be non-redundant in regulating monocyte cell fate, and distinct Notch receptors are likely to act in a concerted manner to coordinate the monocyte differentiation program.

Colony-stimulating factor 1 receptor (CSF1R) signal, CX3CR1, CEBPβ and Nr4a1 have been suggested to be involved in the generation and survival of Ly6Clo monocytes. The lifespan of Ly6Clo monocytes is acutely shortened after blockade of CSF1R, and the numbers of these cells are reduced, and the percentage of dead cells increased in mice with endothelial cell-specific depletion of Csf1(Emoto et al., 2022; MacDonald et al., 2010). CX3CR1 and CEBPβ-knockout mice show accelerated death of Ly6Clo monocyte, display the decreased level of Ly6Clo monocytes(Landsman et al., 2009; Tamura et al., 2017). Nr4a1 knockout mice have significantly fewer monocytes in blood, bone marrow, and spleen, due to differentiation deficiency in MDP and accelerated apoptosis of Ly6Clo monocytes in bone marrow(Hanna et al., 2011). RBP-J deficient mice had normal MDP and Ly6Clo monocytes, expressed normal level of Nr4a1 and Ki-67, exhibited normal half-life and turnover rate, which suggested that RBP-J may not regulate Ly6Clo monocytes through these factors. Further studies are required to elucidate the precise molecular mechanisms of RBP-J in Ly6Clo monocytes under steady state.

At the steady state, Ly6Clo monocytes patrol endothelium of blood vessel. After R848 induced endothelial injury, Ly6Clo monocytes recruit neutrophils to mediate focal endothelial necrosis and fuel vascular inflammation, and subsequently, the Ly6Clo monocytes remove cellular debris(Carlin et al., 2013; Imhof et al., 2016). In response to Listeria monocytogenes infection, Ly6Clo monocytes can extravasate but do not give rise to macrophages or DCs(Auffray et al., 2007). Additional evidence shows that Ly6Clo monocytes do not merely act as luminal blood macrophages. In lung, exposure to unmethylated CpG DNA expands CD16.2+ IM, which originate from Ly6Clo monocytes and spontaneously produce IL-10, thereby preventing allergic inflammation(Schyns et al., 2019). A recent study shows that Ly6Clo monocytes give rise to CD9+ macrophages, which provide an intracellular replication niche for Salmonella Typhimurium in spleen, and Ly6Clo-depleted mice are more resistant to Salmonella Typhimurium infection, suggesting that Ly6Clo monocytes exert certain functions in systemic infection(Hoffman et al., 2021). Our data provide evidence for RBP-J in the function of blood Ly6Clo monocytes, as RBP-J deficient Ly6Clo monocytes exhibited enhanced competition in blood circulation, as well as gave rise to increased numbers of lung CD16.2+ IM. In summary, we showed that RBP-J acted as a fundamental regulator of the maintenance of blood Ly6Clo monocytes and their descendent, at least in part through regulation of CCR2. These results provide insights into understanding the mechanisms that regulate monocyte homeostasis and function.